A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-04-05 analysis/paper/paper.Rmd
update paper

# Introduction
Single-cell -omics technologies now make it possible to model biological systems more accurately than ever before [@tanay_scaling_2017]. One area where single-cell data has been particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation [@etzrodt_quantitativesinglecellapproaches_2014]. Such dynamic processes can be modelled computationally using trajectory inference (TI) methods, which order cells along a trajectory by their overall similarity in -omics datasets (mainly transcriptomics) [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018]. The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also allow to identify more complex trajectory topologies such as cyclic or disconnected graphs. TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process [@tanay_scaling_2017],  thereby allowing the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interactions responsible for one or more bifurcations [@aibar_scenicsinglecellregulatory_2017]. 
Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogues of whole organisms [@regev_scienceforumhuman_2017; @han_mappingmousecell_2018] underline the urgency for accurate, scalable[@aibar_scenicsinglecellregulatory_2017; @angerer_singlecellsmake_2017] and user-friendly TI methods.

<!-- §2: Plethora of new TI methods -->
A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("stable", "implementations")`). For no other single-cell analysis workflows have so many tools been developed, according to several repositories such as omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017]. All methods have their own unique set of characteristics in terms of underlying algorithm, the prior information they require and the outputs they produce. Two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory or not, and what type of topologies they can detect. Early TI methods typically fixed the topology algorithmically (e.g. linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015] or bifurcating [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]), or through parameters provided by the user [@trapnell_dynamicsregulatorscell_2014; @matsumoto_scoupprobabilisticmodel_2016]. These methods therefore mainly focused on correctly ordering the cells along this fixed topology. 
Other methods attempt to infer the topology computationally, which increases the difficulty of the problem at hand, but allows these methods to be applicable on more use cases. Methods that perform topology inference are still in the minority, though current trends suggest this will change soon. A particularly interesting development is presented in the AGA method [@wolf_graphabstractionreconciles_2017] which is to our knowledge the only TI method currently able to deal with disconnected graphs.

@@

<!-- §4: Our study -->
In this study, we performed a comprehensive evaluation of  `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods (`r ref("fig", "ti_evaluation_overview", "a")`, `r ref("sfig", "overview_evaluation_expanded")`). The inclusion criterion for TI methods was primarily based on their free availability and presence of a programming interface (`r ref("stable", "implementations")`). In addition, we added three control methods to define baseline performance: a random trajectory, Component 1 as a control for linear trajectories and Angle as a control for circular trajectories. The evaluation comprised two core aspects: (i) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard on `r sum(tasks_info$task_source == "real")` real and `r sum(tasks_info$task_source == "synthetic")` synthetic datasets, and (ii) a quality control of the provided software and documentation. Finally, we combined both the performance and QC evaluation to provide a set of practical guidelines for users.

@@

<!-- Evaluation overview -->
In order to make gold standard trajectories and predicted trajectories directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "ti_evaluation_overview", "b")`). In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones. The milestone network is automatically classified into one of the nine predefined trajectory types (`r ref("fig", "ti_evaluation_overview", "c")`). We defined a set of metrics for comparing the similarities of such trajectories, each assessing a different aspect of the trajectory: the similarity in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "ti_evaluation_overview", "d")`). The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance. These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`). In order to generate synthetic datasets, we simulated gene regulatory networks using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated single-cell profiling experiments by matching the distributions of the synthetic data with real reference datasets.

@@

<!-- The robustness of the evaluation & development of new methods -->
We found that the performance of a method can be very variable between datasets, and therefore included a large set of both real and synthetic data within our evaluation, leading to a robust overall ranking of the different methods. Nonetheless, we also found that it is possible to define for almost every method a relatively large subset of datasets on which this method on average performs the best. These findings raise some important issues, as we want to avoid that a race to be the first on the ranking would stifle creativity, as there is a high risk that a certain new methodology would not improve upon the state-of-the-art. Indeed, we firmly believe that "good-yet-not-the-best" methods [@norel_selfassessmenttrap_2011] can still provide a very valuable contribution to the field, especially if they make use of novel algorithms, provide a unique insight in specific use cases, or return a more scalable solution. We feel that such methods should also be valued, and encourage their use provided that they are contained in a user friendly tool.

@@

```{r "supplementary", echo=FALSE, results='asis'}
if (params$table_format == "html") 
  source(paste0(dynalysis::get_dynalysis_folder(), "/analysis/paper/sfigs_stables.R"))
write_rds(lst(refs, sfigs, stables), derived_file("sfigs_stables.rds", "paper"))
```

>>>>

# Introduction
Single-cell -omics technologies now make it possible to model biological systems more accurately than ever before [@tanay_scaling_2017]. One area where single-cell data has been particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation [@etzrodt_quantitativesinglecellapproaches_2014]. Such dynamic processes can be modelled computationally using trajectory inference (TI) methods, which order cells along a trajectory by their overall similarity in -omics datasets (mainly transcriptomics) [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018]. The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also allow to identify more complex trajectory topologies such as cyclic [@liu_reconstructingcellcycle_2017] or disconnected [@wolf_graphabstractionreconciles_2017] graphs. TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process [@tanay_scaling_2017],  thereby allowing the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interactions responsible for one or more bifurcations [@aibar_scenicsinglecellregulatory_2017]. 
Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogues of whole organisms [@regev_scienceforumhuman_2017; @han_mappingmousecell_2018] underline the urgency for accurate, scalable[@aibar_scenicsinglecellregulatory_2017; @angerer_singlecellsmake_2017] and user-friendly TI methods.

<!-- §2: Plethora of new TI methods -->
A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("stable", "implementations")`). For no other single-cell analysis workflows have so many tools been developed, according to several repositories such as omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017]. All methods have their own unique set of characteristics in terms of underlying algorithm, the prior information they require and the outputs they produce. Two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory or not, and what type of topologies they can detect. Early TI methods typically fixed the topology algorithmically (e.g. linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015] or bifurcating [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]), or through parameters provided by the user [@trapnell_dynamicsregulatorscell_2014; @matsumoto_scoupprobabilisticmodel_2016]. These methods therefore mainly focused on correctly ordering the cells along the fixed topology. 
Other methods attempt to infer the topology computationally, which increases the difficulty of the problem at hand, but allows these methods to be applicable on more use cases. Methods that perform topology inference are still in the minority, though current trends suggest this will change soon.

@@

<!-- §4: Our study -->
In this study, we performed a comprehensive evaluation (`r ref("fig", "ti_evaluation_overview", "a")`, `r ref("sfig", "overview_evaluation_expanded")`) of  `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods. The inclusion criterion for TI methods was primarily based on their free availability and presence of a programming interface (`r ref("stable", "implementations")`). In addition, we added three control methods to define baseline performance: a random trajectory, Component 1 as a control for linear trajectories and Angle as a control for circular trajectories. The evaluation comprised two core aspects: (i) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard on `r sum(tasks_info$task_source == "real")` real and `r sum(tasks_info$task_source == "synthetic")` synthetic datasets, and (ii) a quality control of the provided software and documentation. Finally, we combined both the performance and QC evaluation to provide a set of practical guidelines for users.

@@

<!-- Evaluation overview -->
In order to make gold standard trajectories and predicted trajectories directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "ti_evaluation_overview", "b")`). In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones. The milestone network is automatically classified into one of the nine predefined trajectory types (`r ref("fig", "ti_evaluation_overview", "c")`). We defined a set of metrics for calculating the similarity of two trajectories represented in this common model, each assessing a different aspect of a trajectory: the cell ordering, the cell neighbourhood and the topology (`r ref("fig", "ti_evaluation_overview", "d")`). The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance. These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`). In order to generate synthetic datasets, we simulated gene regulatory networks using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated single-cell profiling experiments by matching the distributions of the synthetic data with real reference datasets.

@@

<!-- The robustness of the evaluation & development of new methods -->
We found that the performance of a method can be very variable between datasets, and therefore included a large set of both real and synthetic data within our evaluation, leading to a robust overall ranking of the different methods. Nonetheless, we also found that it is possible to define for almost every method a relatively large subset of datasets on which this method on average performs the best. These findings raise some important issues, as we want to avoid that a race to be the first on the ranking would stifle creativity, given that there is a high risk that a certain new methodology would not improve upon the state-of-the-art. Indeed, we firmly believe that "good-yet-not-the-best" methods [@norel_selfassessmenttrap_2011] can still provide a very valuable contribution to the field, especially if they make use of novel algorithms, return a more scalable solution, or provide a unique insight in specific use cases. Some examples for the latter include PhenoPath, which can include additional covariates in its model, Ouija, which returns an measure of uncertainty of each cell's position within the trajectory, and StemID, which can infer the directionality of edges within the trajectory. We feel that such methods should also be valued, and encourage their use provided that they are contained in a user friendly tool.

@@

```{r "supplementary", echo=FALSE, results='asis'}
if (params$table_format == "html") 
  source(paste0(dynalysis::get_dynalysis_folder(), "/analysis/paper/sfigs_stables.R"))
write_rds(lst(refs, sfigs, stables), derived_file("sfigs_stables.rds", "paper"))
write_rds(figs, derived_file("figs.rds", "paper"))
```
